---
title: "<br> <center> Identification of specific gen expression biomarkers of glioma and patient classification with machine learning <br> Omic data analysis"
subtitle: "<center> Bioinformatic and Biostatistic Master Course <br> Universitat Oberta de Catalunya"
author: "<center> Aitor Gómez Osuna"
date: |
  <center> `r format(Sys.Date(),"%e de %B, %Y")`
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    smooth_scroll: true
    collapsed: true
bibliography: bibliography.bib
nocite: |
  @lantz2013machine, @harley15p, @rmarkdowncookbook, @ritchie2015limma,
  @robinson2010edger, @law2014voom, @alexa2009gene, @law2016rna,
  @bioconductor, @reproducibleresearch
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, fig.align = "center" )
```

# Introduction

## Datasets to be used

We selected three datasets (Projects) from the "Harmonized Cancer Datasets" section from the _[Genomic Data Commons Data Portal](https://portal.gdc.cancer.gov/)_ with "transcriptome profiling" as "Data Category" and "RNA-Seq" as "Experimental Strategy":

| __Project__ | __Primary Site__ | __Focus__ | __Usage__ | __Access__ | 
|:------------|:-----------------|:----------|:----------|:-----------|
| TCGA-COAD | colon | Colon Adenocarcinoma | controls | https://portal.gdc.cancer.gov/projects/TCGA-COAD | 
| TCGA-GBM | brain | Glioblastoma Multiforme | cases | https://portal.gdc.cancer.gov/projects/TCGA-GBM |
| TCGA-LGG | brain | Brain Lower Grade Glioma | cases | https://portal.gdc.cancer.gov/projects/TCGA-LGG |

## Planified Analysis

Two main blocks defined the analysis plan:

### Block 1: Differential expression analysis

The 1st block includes the traditional analyses to identify differences in gene expression between cases and controls. Here we will perform two analyses:

1. To identify deferentially expressed (DE) genes between cases and controls.
2. To identify DE genes between the two sets of glioma, since they include samples of different types of glioma.


#### Analysis 1

Here we will compare the glioma sets versus the colon set, one at a time

```
A | TCGA-COAD vs. TCGA-GBM <--- DE genes between colon and glioma
B | TCGA-COAD vs. TCGA-LGG <--- DE genes between colon and glioma
```
#### Analysis 2

From this analysis we want to obtain the DE genes that will help to identify samples from "glioblastoma multiforme" and "brain low grade glioma". Therefore, no "colon adenocarcinoma" samples will be used.

```
C | TCGA-GBM vs. TCGA-LGG <--- DE genes between glioma and low grade glioma
```
  The first analysis helped to identify genes that classify cases and controls. The current result help classify types of cases.

### Block 2: machine learning

This block is focused on machine learning techniques to properly classify patients with (sub-)types of glioma.

#### Analysis 3

This analysis is focused on understanding how clustering algorithms works and classifies samples. Therefore, knn, k-means,  herarchical clustering, Gaussian Mixture Model (GMM) and DBSCAN will be used to classify the samples but:

1. The three datasets with all the features will be clustered (asking for three group in non supervised analysis).  
2. The three datasets but with only the union of the DE genes from analysis 1 and analysis 2.  
3. Only glioma datases with all the features will be clustered (asking for two group in non supervised analysis)  
4. Only glioma datases but with only the features from analysis 2.  


```{r initial packages, message=FALSE, warning=FALSE}
## packages to be used in the analysis
library(Biobase)
library(SummarizedExperiment)
library(TCGAbiolinks)
library(factoextra)
library(edgeR)
library(limma)

library(AnnotationDbi)
library(org.Hs.eg.db)

library(ggvenn)
library(gplots)
library(ggpubr)
library(RColorBrewer)
library(scales)

library(clusterProfiler)
library(rWikiPathways)
library(DOSE)
library(magrittr)
library(enrichplot)
library(Rgraphviz)

library(dplyr)
library(cowplot)
library(kableExtra)
library(tm)
library(wordcloud)
```

# Data

## Data download

First all the data from TGCA projects are download.

```{r tcga_gbm, eval=FALSE}
tcga_gbm <- GDCquery(
  project = "TCGA-GBM",
  data.category = "Transcriptome Profiling",
  experimental.strategy = "RNA-Seq",
  workflow.type = "HTSeq - Counts"
)

GDCdownload(query = tcga_gbm)
tcga_gbm <- GDCprepare(tcga_gbm)
save(tcga_gbm, file="data/tcga_gbm_se.RData")
```

```{r tcga_lgg, eval=FALSE}
tcga_lgg <- GDCquery(
  project = "TCGA-LGG",
  data.category = "Transcriptome Profiling",
  experimental.strategy = "RNA-Seq",
  workflow.type = "HTSeq - Counts"
)

GDCdownload(query = tcga_lgg)
tcga_lgg <- GDCprepare(tcga_lgg)
save(tcga_lgg, file="data/tcga_lgg_se.Rdata")
```

```{r tcga_coad, eval=FALSE}
tcga_coad <- GDCquery(
  project = "TCGA-COAD",
  data.category = "Transcriptome Profiling",
  experimental.strategy = "RNA-Seq",
  workflow.type = "HTSeq - Counts"
)

GDCdownload(query = tcga_coad)
tcga_coad <- GDCprepare(tcga_coad)
save(tcga_coad, file="data/tcga_coad_se.RData")
```

### From SummarizedExperiment to ExpressionSet

The SummarizedExperiment objects are converted to ExpressionSet. Data download of the data is a proccess that may require some time and computational cost. In this sense, data are saved as a Rdata file. 

```{r gbm_exprs, eval=FALSE}
class(tcga_gbm)

tcga_gbm <- ExpressionSet(
    assayDat = assays(tcga_gbm)[[1]],
    phenoData = as(data.frame(colData(tcga_gbm)), "AnnotatedDataFrame"),
    featureData = as(data.frame(rowData(tcga_gbm)), "AnnotatedDataFrame"),
    experimentData = MIAME()
)

class(tcga_gbm)
save(tcga_gbm, file="data/tcga_gbm.RData")
```

```{r lgg_exprs, eval=FALSE}
class(tcga_lgg)

tcga_lgg <- ExpressionSet(
    assayDat = assays(tcga_lgg)[[1]],
    phenoData = as(data.frame(colData(tcga_lgg)), "AnnotatedDataFrame"),
    featureData = as(data.frame(rowData(tcga_lgg)), "AnnotatedDataFrame"),
    experimentData = MIAME()
)

class(tcga_lgg)
save(tcga_lgg, file="data/tcga_lgg.RData")
```

```{r coad_exprs, eval=FALSE}
class(tcga_coad)

tcga_coad <- ExpressionSet(
    assayDat = assays(tcga_coad)[[1]],
    phenoData = as(data.frame(colData(tcga_coad)), "AnnotatedDataFrame"),
    featureData = as(data.frame(rowData(tcga_coad)), "AnnotatedDataFrame"),
    experimentData = MIAME()
)

class(tcga_coad)
save(tcga_coad, file="data/tcga_coad.RData")
```

### Load data

In the case, you need to load the data from file.

```{r load datasets}
load("data/tcga_gbm.RData")
load("data/tcga_lgg.RData")
load("data/tcga_coad.RData")
```

# Analysis 1

## Filtering

The data is generated using RNA-Seq technology and processed with HTSeq, indicating that the expression matrices of the ExpressionSet contains _counts_:

Therefore is should be analyzed using combining `voom` with `limma`.

Let's see the features(genes) in common across the three sets:

```{r common features}
int_fet <- intersect(
  intersect(featureNames(tcga_gbm), featureNames(tcga_lgg)), 
            featureNames(tcga_coad)
)
length(int_fet)
```

```{r join cases}
#Join cases
raw_exprs <- cbind(
  exprs(tcga_gbm[int_fet, ]),
  exprs(tcga_lgg[int_fet, ]),
  exprs(tcga_coad[int_fet, ])
)
dim(raw_exprs)
save(raw_exprs, file="data/raw_exprs.RData")


#groups in dataset
group_f <- as.factor(ifelse(colnames(raw_exprs) %in% sampleNames(tcga_gbm), "gbm",
                          ifelse(colnames(raw_exprs) %in% sampleNames(tcga_lgg), "lgg",
                                 "colon")
                          )
                   )
table(group_f)

#Deleting variables
rm(tcga_coad,tcga_gbm,tcga_lgg, int_fet)
```

In the join dataset, there are `r dim(raw_exprs)[1]` genes and `r dim(raw_exprs)[2]` cases.


### Signal Expression Filtering

Three differents filter approach has been follow:
  + Based on count variance
  + Based on FilterByExprs function from edgeR (CPM equal to 10 divided by median library size in M)
  + Based on standard CPM thresold 1

```{r SEF full data}
#for filtering purpose data are transformed to log-cpm
myCPM <-cpm(raw_exprs)

#mean and median of library sizes in M
col_sum <- colSums(raw_exprs)
mlib <- mean(col_sum) * 1e-6
mlib
medlib <- median(col_sum) * 1e-6
medlib
hist(col_sum * 1e-6, 
     las= 2, col = "navyblue",
     main="Histogram of library sizes", 
     xlab="Library size (M)")
abline(v = mlib, col = "grey")
abline(v = medlib, col = "red")

### Variance filtering
#Based on variance > 1 of counts for each gene
var_set <- apply(raw_exprs, 1, var)
#number of genes retained with var set approach
sum(var_set > 1)


### FilterByExprs function
# Based on FilterByExprs function from edgeR
keep.exprs <- filterByExpr(raw_exprs, group = group_f)
#number of genes retained with var set approach
sum(keep.exprs)


### CPM equal to standard value 1
# threshold to get upper values
thresh <- myCPM > 1
#cpm equal to 1 is a standard threshold

# Genes which are retained are those whose cpm is 
# above thresh in half of the minimum cases of the 
# less abundant group cases
#n samples retained
keepn <- round(min(tabulate(group_f))*0.5,0)
#genes to retain
keepcpm <- rowSums(thresh) >= keepn
#number of genes retained with cpm > 1 approach
sum(keepcpm)
#identifying retained genes
keeprow <- attr(keepcpm,"names")[keepcpm==T]
```

Genes after filtering for every aproach:
  + Based on count variance = `r table(var_set > 1)[2]`  
  + Based on FilterByExprs function from edgeR = `r table(keep.exprs)[2]`
  + Based on standard CPM thresold 1 = `r table(keepcpm)[2]`
  
Between the last two, it shouldn't be a big different, but the first approach is not recommended due to it is calculated only with the variance of the counts, without taking into account library size.

```{r SEF full data 2}
#filter dataset
full_exprs <- raw_exprs[keeprow,]
save(full_exprs, file="data/full_exprs.RData")

#deleting variables
rm(myCPM, var_set, keep.exprs, 
   thresh, keepcpm, keepn, keeprow,
   medlib,mlib,col_sum)
```


## Exploratory analyses

### Principal Component Analysis

As well as how they behave in a PCA:

```{r pca full data, cache = TRUE}
set.seed(123456789)
#### Optional
## Reducing number of cases to reduce computational cost
## If you want to reduce number of cases changes percent
percent <- 1
pca_samples_ids <- sort(sample(x =  1:length(colnames(full_exprs)),
                      size = length(colnames(full_exprs))*percent,
                      replace = FALSE))
pca_group <- group_f[pca_samples_ids]

#Proportion in the full and in the reduced datasets of PCA
prop.table(table(pca_group))
prop.table(table(group_f))

#PCA function prcomp CENTERING but NOT scaling
pca_f <- prcomp(t(full_exprs[,pca_samples_ids]), 
                center = TRUE, 
                scale = FALSE)
#PCA plot
fig_pca.a1 <- fviz_pca_ind(pca_f,
                    geom = "point",
                    col.ind = pca_group
                    )
fig_pca.a1
#deleting variables
rm(pca_f)
```


## Normalization

Normalization is done with edgeR and then voom transformation and calculation of variance weights.

```{r normalization protocol edgeR and limma-voom full data}
countstovoom <- function(data, group, use.method = "TMM", make.plot = TRUE){
  #DGE built object
  dge <- DGEList(data)
  
  #calculations of norm factors based on library size
  normfactor <- calcNormFactors(dge,method = use.method)

  #model design based on dataset groups
  design_mod <- model.matrix(~ 0 + group)
  colnames(design_mod) <- levels(group)

  #voom function object
  return(voom(normfactor,design_mod, plot = make.plot))
  
  #deleting variables
  rm(dge,normfactor)
}

#apply voom fn
v.F <- countstovoom(full_exprs, group_f)
save(v.F, file="data/v.F.RData")
```


### Exploratory analysis after Normalization

### Principal components analysis

How they behave after data normalization in the PCA?

```{r pca normalized full data, cache=TRUE}
#pca function 
pcanorm_f <- prcomp(t(v.F$E[,pca_samples_ids]), 
                    center = TRUE, 
                    scale = FALSE)

#pca plot
fig_pcanorm.a1 <- fviz_pca_ind(pcanorm_f,
                    geom = "point",
                    col.ind = pca_group)
fig_pcanorm.a1

#deleting variables
rm(pcanorm_f, pca_samples_ids, pca_group)
```


## Linear Model 

In this step, lmFit functions built a model using voom data and with makeConstrasts of `limma` groups comparisons are specified.

```{r linear model full data}
#function to built linear model
linmod <- function(voomobj, group, contrast.mod, make.plot = TRUE){
  
  #model design based on dataset groups
  design <- model.matrix(~ 0 + group)
  colnames(design) <- levels(group)
  
  #lmFit function limma
  lm <- lmFit(voomobj,design)
  
  #makeContrasts function
  contmat <- makeContrasts(contrasts = contrast.mod,
                           levels = design)

  #contrasts.fit function based on linear model
  fit.cont <- contrasts.fit(lm,contmat)

  #bayesian transformation
  fit.cont <- eBayes(fit.cont)

  #normalized fitlinear plot
  if (make.plot == TRUE){
    plotSA(fit.cont, main = "voom: Mean-variance trend")
  }
  
  return(fit.cont)
}

#Building linear model 
fit.cont.F <- linmod(v.F,group_f,
                   contrast.mod = c("gbm-colon", "lgg-colon"))
save(fit.cont.F, file="data/fit.cont.F.RData")
```

## DE analysis 1

DEA is performed based on TREAT function and topTREAT [@mccarthy2009testing]. The function `treat` of limma library allow from the fit.cont object and a "log fold change" (logFC) recalculate the moderate t-statistics and p-values. This procedure is much more precise to control FDR than discard genes based on p-values and logFC.

Annotation was done with the `org.Hs.eg.db` database using `ENTREZID`, `SYMBOL` and `GENENAME` features.

```{r DEA 1 full data}
#DEG based on treat function procedure
fit.treat.F <- treat(fit.cont.F,lfc=2)
res.treat.F <- decideTests(fit.treat.F)

#Annotation for fit.treat object
ann <- AnnotationDbi::select(org.Hs.eg.db,
                             keys=rownames(fit.treat.F),
                             keytype= "ENSEMBL",
                             columns=c("ENTREZID","SYMBOL","GENENAME"),
                             multiVals="first")

#removing duplicated in the samples
ann <- ann[!duplicated(ann$ENSEMBL),]

#adding gen labels to fit.treat_genes
fit.treat.F$genes <- ann

#GBM vs colon
topGBM <- topTreat(fit.treat.F,
                   coef= "gbm-colon",
                   sort.by= "logFC",
                   number = sum(summary(
                       res.treat.F)[,"gbm-colon"]))
rownames(topGBM) <- NULL
head(topGBM, 10)


#LGG vs colon
topLGG <- topTreat(fit.treat.F,
                   coef= "lgg-colon",
                   sort.by= "logFC",
                   number = sum(summary(
                       res.treat.F)[,"lgg-colon"]))
rownames(topLGG) <- NULL
head(topLGG, 10)

#savedata
dir.create(path = "DEAsets")
writexl::write_xlsx(topGBM, path = "DEAsets/topGBM.xlsx")
writexl::write_xlsx(topLGG, path = "DEAsets/topLGG.xlsx")
save(topGBM,file = "data/topGBM.RData")
save(topLGG, file = "data/topLGG.RData")

#Sum up of all the DEG for comparisons.
summary(res.treat.F)

#delenting variables
rm(ann, fit.cont.F)
```

### DE genes plots

#### MD plot

Mean-difference plots as a results of the DEA.

```{r MD plots full data}
par(mfrow=c(1,2))
#MD gbm vs colon
md.dea1 <- plotMD(fit.treat.F,coef=1,
       status=res.treat.F[,"gbm-colon"], 
       values=c(-1,1),
       main = "GBM vs colon",
       legend = FALSE,
       cex = 0.1)
abline(h=0,col="grey")

#MD lgg vs colon
plotMD(fit.treat.F,coef=2,
       status=res.treat.F[,"lgg-colon"], 
       values=c(-1,1),
       main = "LGG vs colon",
       legend = FALSE,
       cex = 0.1)
abline(h=0,col="grey")
```

#### Gene sets of DEA1

This chunk consiste on create gene sets for functional analysis afterwards.

```{r Gene sets from full data DEA1}
p.value <- .05
log.FC <- 2
IDgene <- c("ENTREZID", "ENSEMBL","SYMBOL")
parameter <- c("logFC", "adj.P.Val")

##GBM genesets
#universe geneset
gbm.all.genelist <- topGBM[,c(IDgene,parameter)]
gbm.all.gene <- gbm.all.genelist[,"logFC"]
names(gbm.all.gene) <- as.character(gbm.all.genelist[,"ENTREZID"])
gbm.all.gene <- sort(gbm.all.gene, decreasing = TRUE)

#up geneset
gbm.up.genelist <- topGBM[topGBM$adj.P.Val < p.value & topGBM$logFC > log.FC,
                      c(IDgene,parameter)]
gbm.up.gene <- gbm.up.genelist[,"logFC"]
names(gbm.up.gene) <- as.character(gbm.up.genelist[,"ENTREZID"])
gbm.up.gene <- sort(gbm.up.gene, decreasing = TRUE)
#down geneset
gbm.dn.genelist <- topGBM[topGBM$adj.P.Val < p.value & topGBM$logFC < log.FC,
                      c(IDgene,parameter)]
gbm.dn.gene <- gbm.dn.genelist[,"logFC"]
names(gbm.dn.gene) <- as.character(gbm.dn.genelist[,"ENTREZID"])
gbm.dn.gene <- sort(gbm.dn.gene, decreasing = TRUE)

##LGG genesets
#universe geneset
lgg.all.genelist <- topLGG[,c(IDgene,parameter)]
lgg.all.gene <- lgg.all.genelist[,"logFC"]
names(lgg.all.gene) <- as.character(lgg.all.genelist[,"ENTREZID"])
lgg.all.gene <- sort(lgg.all.gene, decreasing = TRUE)

#up geneset
lgg.up.genelist <- topLGG[topLGG$adj.P.Val < p.value & topLGG$logFC > log.FC,
                      c(IDgene,parameter)]
lgg.up.gene <- lgg.up.genelist[,"logFC"]
names(lgg.up.gene) <- as.character(lgg.up.genelist[,"ENTREZID"])
lgg.up.gene <- sort(lgg.up.gene, decreasing = TRUE)

#down geneset
lgg.dn.genelist <- topLGG[topLGG$adj.P.Val < p.value & topLGG$logFC < log.FC,
                      c(IDgene,parameter)]
lgg.dn.gene <- lgg.dn.genelist[,"logFC"]
names(lgg.dn.gene) <- as.character(lgg.dn.genelist[,"ENTREZID"])
lgg.dn.gene <- sort(lgg.dn.gene, decreasing = TRUE)


##function to identify tables after Functional analysis
id.table <- function(tableid, size = 20){
  var.name <<- deparse(substitute(tableid))
  table.name <<- head(tableid, size)
  table.name$table.id <<- rep(var.name,size)  #add a table id
  classif <<- strsplit2(table.name$table.id, split = "\\.")
  colnames(classif) <<- c("tool","cancer","regulation_proccess")
  table.name <- cbind(table.name, classif)
  return(table.name)
}
```

#### Venn Diagram

Venn diagrams are showed as a result of DEA. Genes up and down-regulated are represented for each subtype of glioma.

```{r venn Diagram full data}
#list for venn diagram
venn_list <- list(gbm.up = gbm.up.genelist$ENSEMBL,
                  gbm.down = gbm.dn.genelist$ENSEMBL,
                  lgg.up = lgg.up.genelist$ENSEMBL,
                  lgg.down = lgg.dn.genelist$ENSEMBL)

v.table <- venn(data = venn_list,
                universe = gbm.all.genelist$ENSEMBL,
                small=0.7, showSetLogicLabel=FALSE,
                show.plot=FALSE, intersections=TRUE,
                names = c("GBM vs colon up",
                          "GBM vs colon down",
                          "LGG vs colon up",
                          "LGG vs colon down")
)
isect <- attr(v.table, "intersection")

#venn diagram plot
par(mfrow=c(1,1))
ggvenn(venn_list, 
       fill_color = brewer.pal(8, "Dark2"),
       stroke_size = 0.2, set_name_size = 4.5)
```



###  Specific genes as potential biomarkers of glioma

In this chunk, it is shown genes that are exclusive for GBM vs colon, and LGG vs colon.

```{r specific genes as potential biomarkers of glioma}
#specific genes for up and regulation in GBM excluding common genes with lgg vs colon comparison
gbm.specific <- c(isect[["GBM vs colon up"]],isect[["GBM vs colon down"]])
length(gbm.specific)
gbm.specific.table <- topGBM[topGBM$ENSEMBL %in% gbm.specific,]
head(gbm.specific.table)

#specific genes for up and down regulation in LGG excluding common genes with gbm vs colon comparison
lgg.specific <- c(isect[["LGG vs colon up"]],isect[["LGG vs colon down"]])
length(lgg.specific)
lgg.specific.table <- topLGG[topLGG$ENSEMBL %in% lgg.specific,]
head(lgg.specific.table)
```


### Functional analysis DEA1

#### rWikipatways functional analysis

Analysis from rwikipathways [@slenter2018wikipathways].

clusterProfiler provides `enricher` function for hypergeometric test and `GSEA` function for gene set enrichment analysis that are designed to accept user defined annotation. First we are gonna update information for Wikipathways and download the data from database.

```{r wikipathways data download and preparation DEA1}
#Download pathways from homo sapiens to localfile
wp.hs.gmt<- downloadPathwayArchive(organism="Homo sapiens",
                       format="gmt")
wp2gene <- readPathwayGMT(wp.hs.gmt)

#Establish ID for wikipathways database
wpid2gene <- wp2gene %>% dplyr::select(wpid,gene) #TERM2GENE
wpid2name <- wp2gene %>% dplyr::select(wpid,name) #TERM2NAME
```


##### ORA analysis just needs a list of ENTREZ ids.

```{r ORA in rwikipathways DEA1}
### GBM 
#ORA pathway up
ewp.gbm.up <- enricher(gene = names(gbm.up.gene),
                       universe = names(gbm.all.gene),
                       pAdjustMethod = "BH",
                       pvalueCutoff = .05,
                       TERM2GENE = wpid2gene,
                       TERM2NAME = wpid2name)
ewp.gbm.up <- DOSE::setReadable(ewp.gbm.up, org.Hs.eg.db, keyType = "ENTREZID")
head(ewp.gbm.up)
p1<-barplot(ewp.gbm.up, 
            showCategory = 7, 
            font.size = 7,
            label_format = 15)+
  ggtitle("GBM vs Colon DEG up")
ewp.gbm.up <- id.table(ewp.gbm.up)



#ORA pathway dn
ewp.gbm.dn <- enricher(gene = names(gbm.dn.gene),
                       universe = names(gbm.all.gene),
                       pAdjustMethod = "BH",
                       pvalueCutoff = .05,
                       TERM2GENE = wpid2gene,
                       TERM2NAME = wpid2name)
ewp.gbm.dn <- DOSE::setReadable(ewp.gbm.dn, org.Hs.eg.db, keyType = "ENTREZID")
head(ewp.gbm.dn)
p2<-barplot(ewp.gbm.dn,            
            showCategory = 7, 
            font.size = 7,
            label_format = 15)+
  ggtitle("GBM vs Colon DEG down")
ewp.gbm.dn <- id.table(ewp.gbm.dn, size = 12)

### LGG 
#ORA pathway up
ewp.lgg.up <- enricher(gene = names(lgg.up.gene),
                       universe = names(lgg.all.gene),
                       pAdjustMethod = "BH",
                       pvalueCutoff = .05,
                       TERM2GENE = wpid2gene,
                       TERM2NAME = wpid2name)
ewp.lgg.up <- DOSE::setReadable(ewp.lgg.up, org.Hs.eg.db, keyType = "ENTREZID")
head(ewp.lgg.up)
p3<-barplot(ewp.lgg.up,             
            showCategory = 7, 
            font.size = 7,
            label_format = 15) +
  ggtitle("LGG vs Colon DEG up")
ewp.lgg.up <- id.table(ewp.lgg.up)

#ORA pathway dn
ewp.lgg.dn <- enricher(gene = names(lgg.dn.gene),
                       universe = names(lgg.all.gene),
                       pAdjustMethod = "BH",
                       pvalueCutoff = .05,
                       TERM2GENE = wpid2gene,
                       TERM2NAME = wpid2name)
ewp.lgg.dn <- DOSE::setReadable(ewp.lgg.dn, org.Hs.eg.db, keyType = "ENTREZID")
head(ewp.lgg.dn)
p4<-barplot(ewp.lgg.dn,            
            showCategory = 7, 
            font.size = 7,
            label_format = 15)+
  ggtitle("LGG vs Colon DEG down")
ewp.lgg.dn <- id.table(ewp.lgg.dn)

#plot grid for results
pt <- plot_grid(p1,p2,p3,p4, ncol = 2)

## to collect all functional analysis results
keep.colums.enricher <- c("ID","Description","pvalue","p.adjust","Count","geneID",
                          "table.id","tool","cancer","regulation_proccess")
fa.DEA1 <- rbind(ewp.gbm.up[,keep.colums.enricher], 
                  ewp.gbm.dn[,keep.colums.enricher], 
                  ewp.lgg.up[,keep.colums.enricher], 
                  ewp.lgg.dn[,keep.colums.enricher])
```

##### GSEA analysis with wikipathways

```{r GSEA in rwikpathways DEA1}
set.seed(1234567)
## GBM GSEA
gwp.gbm <- GSEA(geneList = c(gbm.up.gene,
                             gbm.dn.gene),
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                TERM2GENE = wpid2gene,
                TERM2NAME = wpid2name,
                verbose = FALSE
                   )
gwp.gbm <- setReadable(gwp.gbm, 'org.Hs.eg.db', 'ENTREZID')

#network plot gbm gsea
cnetplot(gwp.gbm, 
         foldChange=c(gbm.up.gene,
                      gbm.dn.gene),
         repel = TRUE,
         showCategory = 15,
         colorEdge = FALSE,
         circular = FALSE,
         cex_category = 0.6,
         cex_gene = 0.4,
         cex_label_category = 0.6,
         cex_label_gene = 0.3,
         main = "GBM vs colon")

#gbm gsea wikipathways up
gwp.gbm.up <- gwp.gbm[which(gwp.gbm$NES > 1),]
head(gwp.gbm.up)
gwp.gbm.up <- id.table(gwp.gbm.up, size = 17)

#gbm gsea wikipathways down
gwp.gbm.dn <- gwp.gbm[which(gwp.gbm$NES < -1),]
head(gwp.gbm.dn)
gwp.gbm.dn <- id.table(gwp.gbm.dn, size = 5)



## LGG GSEA
gwp.lgg <- GSEA(geneList = c(lgg.up.gene,
                             lgg.dn.gene),
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                TERM2GENE = wpid2gene,
                TERM2NAME = wpid2name,
                verbose = FALSE
                   )
gwp.lgg <- setReadable(gwp.lgg, 'org.Hs.eg.db', 'ENTREZID')

#network plot lgg gsea
cnetplot(gwp.lgg, 
         categorySize="pvalue", 
         foldChange=c(lgg.up.gene,
                      lgg.dn.gene),
         repel = TRUE,
         showCategory = 15,
         colorEdge = FALSE,
         circular = FALSE,
         cex_category = 0.6,
         cex_gene = 0.4,
         cex_label_category = 0.6,
         cex_label_gene = 0.3)

#lgg gsea wikipathways up
gwp.lgg.up <- gwp.lgg[which(gwp.lgg$NES > 1),]
head(gwp.lgg.up)
gwp.lgg.up <- id.table(gwp.lgg.up)
#lgg gsea wikipathways down
gwp.lgg.dn <- gwp.lgg[which(gwp.lgg$NES < -1),]
head(gwp.lgg.dn)
gwp.lgg.dn <- id.table(gwp.lgg.dn)

## to collect all functional analysis results
keep.colums.gsea <- c("ID","Description","pvalue","p.adjust","setSize","core_enrichment",
                      "table.id","tool","cancer","regulation_proccess")
fa.DEA1 <- rbind(fa.DEA1,
                  setNames(gwp.gbm.up[,keep.colums.gsea], names(fa.DEA1)),
                  setNames(gwp.gbm.dn[,keep.colums.gsea],names(fa.DEA1)),
                  setNames(gwp.lgg.up[,keep.colums.gsea],names(fa.DEA1)),
                  setNames(gwp.lgg.dn[,keep.colums.gsea],names(fa.DEA1)))
```

#### Gene Ontology (GO) functional analysis

Gene ontology functional analysis with `clusterProfiler` functions `enrichGo` adn `gseGO`.
Due to the parent-childhood annotation of GO terms, `dropGO` is applied to avoid general terms, eliminating higher levels of GO.
Only biology proccess anotation is set for the analysis.

##### ORA in GO

```{r ORA in GO DEA1}
#level to drop
level.go <- 5

### GBM
## BIOLOGICAL PROCESS (BP)
# ORA GO up BP
ego.gbm.up <- enrichGO(gene = gbm.up.genelist$ENSEMBL,
                          OrgDb = org.Hs.eg.db,
                          keyType = 'ENSEMBL',
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.01,
                          qvalueCutoff  = 0.05,
                          readable = TRUE)

ego.gbm.up <- setReadable(dropGO(ego.gbm.up,level = 1:level.go),
                            org.Hs.eg.db, keyType = "ENTREZID")
head(ego.gbm.up)
ego.gbm.up <- id.table(ego.gbm.up)

# ORA GO down BP
ego.gbm.dn <- enrichGO(gene = gbm.dn.genelist$ENSEMBL,
                          OrgDb = org.Hs.eg.db,
                          keyType = 'ENSEMBL',
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.01,
                          qvalueCutoff  = 0.05,
                          readable = TRUE)
ego.gbm.dn <- setReadable(dropGO(ego.gbm.dn,level = 1:level.go),
                            org.Hs.eg.db, keyType = "ENTREZID")
head(ego.gbm.dn)
ego.gbm.dn <- id.table(ego.gbm.dn)

### LGG
## BIOLOGICAL PROCESS (BP)
# ORA GO up BP
ego.lgg.up <- enrichGO(gene = lgg.up.genelist$ENSEMBL,
                          OrgDb = org.Hs.eg.db,
                          keyType = 'ENSEMBL',
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.01,
                          qvalueCutoff  = 0.05,
                          readable = TRUE)

ego.lgg.up <- setReadable(dropGO(ego.lgg.up,level = 1:level.go),
                            org.Hs.eg.db, keyType = "ENTREZID")
head(ego.lgg.up)
ego.lgg.up <- id.table(ego.lgg.up)

# ORA GO down BP
ego.lgg.dn <- enrichGO(gene = lgg.dn.genelist$ENSEMBL,
                          OrgDb = org.Hs.eg.db,
                          keyType = 'ENSEMBL',
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.01,
                          qvalueCutoff  = 0.05,
                          readable = TRUE)
ego.lgg.dn <- setReadable(dropGO(ego.lgg.dn,level = 1:level.go),
                            org.Hs.eg.db, keyType = "ENTREZID")
head(ego.lgg.dn)
ego.lgg.dn <- id.table(ego.lgg.dn)



## to collect all functional analysis results 
fa.DEA1 <- rbind(fa.DEA1,
                  ego.gbm.up[,keep.colums.enricher], 
                  ego.gbm.dn[,keep.colums.enricher], 
                  ego.lgg.up[,keep.colums.enricher], 
                  ego.lgg.dn[,keep.colums.enricher]) 
```


##### GSEA in Gene Ontology


```{r GSEA in GO DEA1, message=FALSE}
set.seed(1234567)

### GBM
## BIOLOGICAL PROCESS (BP)
# GSEA GO BP
ggo.gbm.BP <- gseGO(geneList = c(gbm.up.gene,
                                 gbm.dn.gene),
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pvalueCutoff = 0.05,
                    eps = 0,
                    verbose = FALSE)
ggo.gbm.BP <- setReadable(ggo.gbm.BP,'org.Hs.eg.db', 'ENTREZID')

#gbm gsea GO up
ggo.gbm.up <- ggo.gbm.BP[which(ggo.gbm.BP$NES > 1),]
head(ggo.gbm.up)
ggo.gbm.up <- id.table(ggo.gbm.up)

#gbm gsea GO down
ggo.gbm.dn <- ggo.gbm.BP[which(ggo.gbm.BP$NES < -1),]
head(ggo.gbm.dn)
ggo.gbm.dn <- id.table(ggo.gbm.dn)

### LGG
## BIOLOGICAL PROCESS (BP)
# GSEA GO BP
ggo.lgg.BP <- gseGO(geneList = c(lgg.up.gene,
                                 lgg.dn.gene),
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pvalueCutoff = 0.05,
                    eps = 0,
                    verbose = FALSE)
ggo.lgg.BP <- setReadable(ggo.lgg.BP, 'org.Hs.eg.db', 'ENTREZID')

#lgg gsea GO up
ggo.lgg.up <- ggo.lgg.BP[which(ggo.lgg.BP$NES > 1),]
head(ggo.lgg.up)
ggo.lgg.up <- id.table(ggo.lgg.up)

#lgg gsea GO down
ggo.lgg.dn <- ggo.lgg.BP[which(ggo.lgg.BP$NES < -1),]
head(ggo.lgg.dn)
ggo.lgg.dn <- id.table(ggo.lgg.dn)

## to collect all functional analysis results
fa.DEA1 <- rbind(fa.DEA1,
                  setNames(ggo.gbm.up[,keep.colums.gsea], names(fa.DEA1)),
                  setNames(ggo.gbm.dn[,keep.colums.gsea],names(fa.DEA1)),
                  setNames(ggo.lgg.up[,keep.colums.gsea],names(fa.DEA1)),
                  setNames(ggo.lgg.dn[,keep.colums.gsea],names(fa.DEA1)))
```

#### Kyoto Encyclopedia of Genes and Genomes (KEGG) functional analysis

Similar procedure was done with the KEGG database. In this case the functions are `enrichKEGG` and `gseKEGG` for ORA and GSEA analysis, respectively.

##### ORA in KEGG

```{r ORA in KEGG DEA1}
### GBM
# ORA KEGG up
ekg.gbm.up <- enrichKEGG(gene = names(gbm.up.gene),
                         organism = "hsa",
                         pAdjustMethod = "BH",
                         universe = names(gbm.all.gene),
                         pvalueCutoff = 0.05)

ekg.gbm.up <- setReadable(ekg.gbm.up, org.Hs.eg.db, keyType = "ENTREZID")
head(ekg.gbm.up)
ekg.gbm.up <- id.table(ekg.gbm.up)

# ORA KEGG down
ekg.gbm.dn <- enrichKEGG(gene = names(gbm.dn.gene),
                         organism = "hsa",
                         pAdjustMethod = "BH",
                         universe = names(gbm.all.gene),
                         pvalueCutoff = 0.05)

ekg.gbm.dn <- setReadable(ekg.gbm.dn, org.Hs.eg.db, keyType = "ENTREZID")
head(ekg.gbm.dn)
ekg.gbm.dn <- id.table(ekg.gbm.dn)

### LGG
# ORA KEGG up
ekg.lgg.up <- enrichKEGG(gene = names(lgg.up.gene),
                         organism = "hsa",
                         pAdjustMethod = "BH",
                         universe = names(lgg.all.gene),
                         pvalueCutoff = 0.05)

ekg.lgg.up <- setReadable(ekg.lgg.up, org.Hs.eg.db, keyType = "ENTREZID")
head(ekg.lgg.up)
ekg.lgg.up <- id.table(ekg.lgg.up)

# ORA KEGG down
ekg.lgg.dn <- enrichKEGG(gene = names(lgg.dn.gene),
                         organism = "hsa",
                         pAdjustMethod = "BH",
                         universe = names(lgg.all.gene),
                         pvalueCutoff = 0.05)

ekg.lgg.dn <- setReadable(ekg.lgg.dn, org.Hs.eg.db, keyType = "ENTREZID")
head(ekg.lgg.dn)
ekg.lgg.dn <- id.table(ekg.lgg.dn)


## to collect all functional analysis results
fa.DEA1 <- rbind(fa.DEA1,
                  ekg.gbm.up[,keep.colums.enricher], 
                  ekg.gbm.dn[,keep.colums.enricher], 
                  ekg.lgg.up[,keep.colums.enricher], 
                  ekg.lgg.dn[,keep.colums.enricher])
```


##### GSEA in KEGG


```{r GSEA in KEGG DEA1}
set.seed(1234567)

##GBM
#GSEA in KEGG
gkg.gbm <- gseKEGG(geneList = c(gbm.up.gene,
                                gbm.dn.gene),
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   verbose = FALSE
                   )
gkg.gbm <- setReadable(gkg.gbm, 'org.Hs.eg.db', 'ENTREZID')

#gbm gsea kegg up
gkg.gbm.up <- gkg.gbm[which(gkg.gbm$NES > 1),]
head(gkg.gbm.up)
gkg.gbm.up <- id.table(gkg.gbm.up)

#gbm gsea kegg down
gkg.gbm.dn <- gkg.gbm[which(gkg.gbm$NES < -1),]
head(gkg.gbm.dn)
gkg.gbm.dn <- id.table(gkg.gbm.dn)

##lgg
#GSEA in KEGG
gkg.lgg <- gseKEGG(geneList = c(lgg.up.gene,
                                lgg.dn.gene),
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   verbose = FALSE
                   )
gkg.lgg <- setReadable(gkg.lgg, 'org.Hs.eg.db', 'ENTREZID')

#lgg gsea kegg up
gkg.lgg.up <- gkg.lgg[which(gkg.lgg$NES > 1),]
head(gkg.lgg.up)
gkg.lgg.up <- id.table(gkg.lgg.up)

#lgg gsea kegg down
gkg.lgg.dn <- gkg.lgg[which(gkg.lgg$NES < -1),]
head(gkg.lgg.dn)
gkg.lgg.dn <- id.table(gkg.lgg.dn)

fa.DEA1 <- rbind(fa.DEA1,
                  setNames(gkg.gbm.up[,keep.colums.gsea],names(fa.DEA1)),
                  setNames(gkg.gbm.dn[,keep.colums.gsea],names(fa.DEA1)),
                  setNames(gkg.lgg.up[,keep.colums.gsea],names(fa.DEA1)),
                  setNames(gkg.lgg.dn[,keep.colums.gsea],names(fa.DEA1)))
writexl::write_xlsx(fa.DEA1, path = "DEAsets/fa.DEA1.xlsx")
```


#### Wordcloud functional analysis from DEA1

In this section, different wordclouds are shown from proccesses and genes involved in the proccesses.

```{r wordclouds FA1}
### fuction to create tables and wordclouds of proccess and genes respectively
proccess.table <- function(fa.table, database, cases, reg){
  t <- fa.table
  enrich <- paste("e",database, sep = "")
  gsea <- paste("g",database, sep = "")
  t <- t$Description[which(t$tool == enrich & t$cancer == cases & t$regulation_proccess == reg|
                            t$tool == gsea & t$cancer == cases & t$regulation_proccess == reg)]
  t <- as.data.frame(table(t))
  t <- t[order(t$Freq,decreasing = TRUE),]
  return <- (t)
}

#function to get wordcloud
getwordcloud <- function(fa.table, database, cases, reg,  scale.w = c(1.8,.3)){
  w <- fa.table
  enrich <- paste("e",database, sep = "")
  gsea <- paste("g",database, sep = "")
  w <- w$geneID[which(w$tool == enrich & w$cancer == cases & w$regulation_proccess == reg|
                        w$tool == gsea & w$cancer == cases & w$regulation_proccess == reg)]
  w <- unlist(strsplit(w, "/"))
  w <- as.data.frame(table(w))
  wordcloud(words = w$w, freq = w$Freq,
            min.freq = 4,max.words=50, random.order=FALSE, 
            rot.per=0.35, scale= scale.w, colors=brewer.pal(8, "Dark2"))
}

#### Wordclouds for FA1 ####
### acronyms
## databases
# wp = wikipathways
# go = gene ontology
# kg = KEGG
##cases
# gbm = GBM vs colon
# lgg = LGG vs colon
# gli = GBM vs LGG

### WIKIPATHWAYS ####
## GBM 
#up-regulated
fa.wp.gbm.up<- proccess.table(fa.table = fa.DEA1, 
                           database = "wp", 
                           cases = "gbm", 
                           reg = "up")
head(fa.wp.gbm.up)
getwordcloud(fa.table = fa.DEA1, 
             database = "wp",
             cases = "gbm", 
             reg = "up")
#down-regulated
fa.wp.gbm.dn<- proccess.table(fa.table = fa.DEA1, 
                           database = "wp", 
                           cases = "gbm", 
                           reg = "dn")
head(fa.wp.gbm.dn)
getwordcloud(fa.table = fa.DEA1, 
             database = "wp",
             cases = "gbm", 
             reg = "dn")

## LGG
#up
fa.wp.lgg.up<- proccess.table(fa.table = fa.DEA1, 
                           database = "wp", 
                           cases = "lgg", 
                           reg = "up")
head(fa.wp.lgg.up)
getwordcloud(fa.table = fa.DEA1, 
             database = "wp",
             cases = "lgg", 
             reg = "up")
#down-regulated
fa.wp.lgg.dn<- proccess.table(fa.table = fa.DEA1, 
                           database = "wp", 
                           cases = "lgg", 
                           reg = "dn")
head(fa.wp.lgg.dn)
getwordcloud(fa.table = fa.DEA1, 
             database = "wp",
             cases = "lgg", 
             reg = "dn")


## GENE ONTOLOGY ####
## GBM 
#up-regulated
fa.go.gbm.up<- proccess.table(fa.table = fa.DEA1, 
                           database = "go", 
                           cases = "gbm", 
                           reg = "up")
head(fa.go.gbm.up)
getwordcloud(fa.table = fa.DEA1, 
             database = "go",
             cases = "gbm", 
             reg = "up")
#down-regulated
fa.go.gbm.dn<- proccess.table(fa.table = fa.DEA1, 
                           database = "go", 
                           cases = "gbm", 
                           reg = "dn")
head(fa.go.gbm.dn)
getwordcloud(fa.table = fa.DEA1, 
             database = "go",
             cases = "gbm", 
             reg = "dn",
             scale.w = c(1,.2))

## LGG
#up
fa.go.lgg.up<- proccess.table(fa.table = fa.DEA1, 
                           database = "go", 
                           cases = "lgg", 
                           reg = "up")
head(fa.go.lgg.up)
getwordcloud(fa.table = fa.DEA1, 
             database = "go",
             cases = "lgg", 
             reg = "up")
#down-regulated
fa.go.lgg.dn<- proccess.table(fa.table = fa.DEA1, 
                           database = "go", 
                           cases = "lgg", 
                           reg = "dn")
head(fa.go.lgg.dn)
getwordcloud(fa.table = fa.DEA1, 
             database = "go",
             cases = "lgg", 
             reg = "dn",
             scale.w = c(1,.2))

## KEGG ####
## GBM 
#up-regulated
fa.kg.gbm.up<- proccess.table(fa.table = fa.DEA1, 
                           database = "kg", 
                           cases = "gbm", 
                           reg = "up")
head(fa.kg.gbm.up)
getwordcloud(fa.table = fa.DEA1, 
             database = "kg",
             cases = "gbm", 
             reg = "up")
#down-regulated
fa.kg.gbm.dn<- proccess.table(fa.table = fa.DEA1, 
                           database = "kg", 
                           cases = "gbm", 
                           reg = "dn")
head(fa.kg.gbm.dn)
getwordcloud(fa.table = fa.DEA1, 
             database = "kg",
             cases = "gbm", 
             reg = "dn")

## LGG
#up
fa.kg.lgg.up<- proccess.table(fa.table = fa.DEA1, 
                           database = "kg", 
                           cases = "lgg", 
                           reg = "up")
head(fa.kg.lgg.up)
getwordcloud(fa.table = fa.DEA1, 
             database = "kg",
             cases = "lgg", 
             reg = "up")
#down-regulated
fa.kg.lgg.dn<- proccess.table(fa.table = fa.DEA1, 
                           database = "kg", 
                           cases = "lgg", 
                           reg = "dn")
head(fa.kg.lgg.dn)
getwordcloud(fa.table = fa.DEA1, 
             database = "kg",
             cases = "lgg", 
             reg = "dn")
```


# Analysis 2

In the second analysis, we want to carry out an differential expression analysis between similar glioma cancer typology. Then, no colon cases will be use below.

## Filtering

Colon cases were removed and new factor group was created in this chunk.

```{r sim dataset}
#sim data, no colon cases
sim_exprs <- raw_exprs[,which(group_f != "colon")]

#groups for sim data
group_s <- as.factor(as.character(
    group_f[which(group_f != "colon")]))
```

### Signal Expression Filtering

Similar procedure is used to do the signal expression filtering. CPM > 1 sets was employed.

```{r SEF sim data}
#for filtering purpose data are transformed to log-cpm
myCPM <-cpm(sim_exprs)

#mean and median of library sizes in M
col_sum <- colSums(sim_exprs)
mlib <- mean(col_sum) * 1e-6
mlib
medlib <- median(col_sum) * 1e-6
medlib
hist(col_sum * 1e-6, 
     las= 2, col = "navyblue",
     main="Histogram of library sizes analysis 2", 
     xlab="Library size (M)")
abline(v = mlib, col = "grey")
abline(v = medlib, col = "red")

### CPM equal to standard value 1
# threshold to get upper values
thresh <- myCPM > 1
#cpm equal to 1 is a standard threshold

# Genes which are retained are those whose cpm is 
# above thresh in half of the minimum cases of the 
# less abundant group cases
#n samples retained
keepn <- round(min(tabulate(group_f))*0.5,0)
#genes to retain
keepcpm <- rowSums(thresh) >= keepn
#number of genes retained with cpm > 1 approach
sum(keepcpm)
#identifying retained genes
keeprow <- attr(keepcpm,"names")[keepcpm==T]
```



```{r SEF sim data2}
#filter dataset
sim_exprs <- sim_exprs[keepcpm,]
save(sim_exprs, file="data/sim_exprs.RData")

#deleting variables
rm(myCPM, thresh, keepcpm, keepn, keeprow, 
   medlib,mlib,col_sum)
```


## Exploratory analyses

#### Principal Component Analysis

Again, as well as how they behave in a PCA:

```{r pca sim data, cache=TRUE}
set.seed(1234567)

#### Optional
## Reducing number of cases to reduce computational cost
## If you want to reduce number of cases changes percent
percent <- 1
pca_samples_ids <- sort(sample(x =  1:length(colnames(sim_exprs)),
                      size = length(colnames(sim_exprs))*percent,
                      replace = FALSE))
pca_group <- group_s[pca_samples_ids]

#Proportion in the full and in the reduced datasets of PCA
prop.table(table(pca_group))
prop.table(table(group_s))

#pca
pca_s <- prcomp(t(sim_exprs[,pca_samples_ids]), 
                center = TRUE, 
                scale = FALSE)

fig_pca.a2 <- fviz_pca_ind(pca_s,
                    geom = "point",
                    col.ind = pca_group
                    )
fig_pca.a2

#deleting variables
rm(pca_s)
```

## Normalization

Normalization with edgeR and voom, as previously done in analysis 1.

```{r DGE list sim data}
#apply countstovoom function
v.S <- countstovoom(sim_exprs, group_s)
save(v.S, file="data/v.S.RData")
```


### Exploratory analysis after Normalization

#### Principal components analysis

How are they after normalization proccess? 

```{r pca normalized sim data}
#pca function 
pcanorm_s <- prcomp(t(v.S$E[,pca_samples_ids]), 
                    center = TRUE, 
                    scale = FALSE)

#pca plot
fig_pcanorm.a2 <- fviz_pca_ind(pcanorm_s,
                    geom = "point",
                    col.ind = pca_group)
fig_pcanorm.a2

#deleting variables
rm(pcanorm_s, pca_group, pca_samples_ids)
```

## Linear Model

Again, `linmod` to create the linear model.

```{r Linear model sim data}
#Building linear model 
fit.cont.S <- linmod(voomobj = v.S,
                   group = group_s,
                   contrast.mod = as.character(c("gbm-lgg"))
                   )
save(fit.cont.S, file="data/fit.cont.S.RData")
```

## DE analysis 2

In this second DE analysis GBM is compared with LGG. So reference levels are taken for LGG and GBM face to this gene-expression levels.

```{r DEA 2 sim data}
#DEG based on treat function procedure
fit.treat.S <- treat(fit.cont.S,lfc=2)
res.treat.S <- decideTests(fit.treat.S)

#Annotation for fit.treat object
ann <- AnnotationDbi::select(org.Hs.eg.db,
                             keys=rownames(fit.treat.S),
                             keytype= "ENSEMBL",
                             columns=c("ENTREZID","SYMBOL","GENENAME"),
                             multiVals="first")

#removing duplicated in the samples
ann <- ann[!duplicated(ann$ENSEMBL),]

#adding gen labels to fit.treat_genes
fit.treat.S$genes <- ann

#GBM vs LGG
topGLIOMA <- topTreat(fit.treat.S,
                   coef= "gbm-lgg",
                   sort.by= "logFC",
                   number = sum(summary(
                       res.treat.S)[,"gbm-lgg"]))
rownames(topGLIOMA) <- NULL

#potential biomarkers of GBM
head(topGLIOMA, 10)

#savedata
writexl::write_xlsx(topGLIOMA, path = "DEAsets/topGLIOMA.xlsx")
save(topGLIOMA, file = "data/topGLIOMA.RData")

#Sum up of all the DEG for comparisons.
summary(res.treat.S)

#deleting variables
rm(ann, fit.cont.S)
```

### DE genes plots

#### MD plot

Mean-difference plots as a results of the DEA.

```{r MD plots sim data}
#MD gbm vs lgg
plotMD(fit.treat.S,coef=1,
       status=res.treat.S[,"gbm-lgg"], 
       values=c(-1,1),
       main = "GBM vs LGG",
       cex = 0.1,
       legend = FALSE)
  abline(h=0,col="grey")
```


#### Gene sets of DEA2

This chunk consiste on create gene sets for functional analysis afterwards.

```{r Gene sets from sim data DEA2}
p.value <- .05
log.FC <- 2
IDgene <- c("ENTREZID", "ENSEMBL","SYMBOL")
parameter <- c("logFC", "adj.P.Val")

##GLIOMA genesets
#universe geneset
gli.all.genelist <- topGLIOMA[,c(IDgene,parameter)]
gli.all.gene <- gli.all.genelist[,"logFC"]
names(gli.all.gene) <- as.character(gli.all.genelist[,"ENTREZID"])
gli.all.gene <- sort(gli.all.gene, decreasing = TRUE)

#up geneset
gli.up.genelist <- topGLIOMA[topGLIOMA$adj.P.Val < p.value & topGLIOMA$logFC > log.FC,
                      c(IDgene,parameter)]
gli.up.gene <- gli.up.genelist[,"logFC"]
names(gli.up.gene) <- as.character(gli.up.genelist[,"ENTREZID"])
gli.up.gene <- sort(gli.up.gene, decreasing = TRUE)
#down geneset
gli.dn.genelist <- topGLIOMA[topGLIOMA$adj.P.Val < p.value & topGLIOMA$logFC < log.FC,
                      c(IDgene,parameter)]
gli.dn.gene <- gli.dn.genelist[,"logFC"]
names(gli.dn.gene) <- as.character(gli.dn.genelist[,"ENTREZID"])
gli.dn.gene <- sort(gli.dn.gene, decreasing = TRUE)
```

#### Venn Diagram

Venn diagram plot in this second analysis was done with all of the result from the analysis 1 and with the results of the second analysis. Again up and down-regulated genes are separated.

```{r venn Diagram sim data}
#list for venn diagram
venn_list2 <- list(a1.up = unique(c(gbm.up.genelist$ENSEMBL,
                                    lgg.up.genelist$ENSEMBL)),
                   a1.dn = unique(c(gbm.dn.genelist$ENSEMBL,
                                    lgg.dn.genelist$ENSEMBL)),
                   gli.up = gli.up.genelist$ENSEMBL,
                   gli.dn = gli.dn.genelist$ENSEMBL)

v.table2 <- venn(data = venn_list2,
                universe = gli.all.genelist$ENSEMBL,
                small=0.7, showSetLogicLabel=FALSE,
                show.plot=FALSE, intersections=TRUE,
                names = c("analysis 1 up",
                          "analysis 1 down",
                          "GBM vs LGG up",
                          "GBM vs LGG down")
)
isect2 <- attr(v.table2, "intersection")

par(mfrow=c(1,1))
#venn diagram
ggvenn(venn_list2, 
       fill_color = brewer.pal(8, "Dark2"),
       stroke_size = 0.2, set_name_size = 4.5,
       text_size = 3.5)
```


### Specific genes as potential biomarkers of GBM and LGG

Previous `TOPGLIOMA` dataset can be taken as a list of potential biomarkers but in this section we explore the groups represented in the Venn diagram which are exclusive for some reason.

```{r specific genes as potential biomarkers of GBM and LGG}
#specific genes for up and down regulation in GBM excluding common genes with analysis 1
gbm.specific2 <- c(isect2[["GBM vs LGG up"]],isect2[["GBM vs LGG down"]])
length(gbm.specific2)
gbm.specific.table2 <- topGLIOMA[topGLIOMA$ENSEMBL %in% gbm.specific2,]
head(gbm.specific.table2)

#specific genes for up and down regulation in GBM and up and down analysis 1, respectively
gbm.equal.a1 <- c(isect2[["analysis 1 up:GBM vs LGG up"]],isect2[["analysis 1 down:GBM vs LGG down"]])
length(gbm.equal.a1)
gbm.equal.a1.table <- topGLIOMA[topGLIOMA$ENSEMBL %in% gbm.equal.a1,]
head(gbm.equal.a1.table)

#specific genes for up and down regulation in GBM and down and up analysis 1, respectively
gbm.vic.a1 <- c(isect2[["analysis 1 down:GBM vs LGG up"]],isect2[["analysis 1 up:GBM vs LGG down"]])
length(gbm.vic.a1)
gbm.vic.a1.table <- topGLIOMA[topGLIOMA$ENSEMBL %in% gbm.vic.a1,]
head(gbm.vic.a1.table)
```

### Functional analysis DEA2

Similar procedure as functional analysis in DEA1

#### rWikipatways functional analysis


##### ORA analysis with wikipathways


```{r ORA in rwikipathways DEA2}
### GBM vs LGG 
#ORA pathway up
ewp.gli.up <- enricher(gene = names(gli.up.gene),
                       universe = names(gli.all.gene),
                       pAdjustMethod = "BH",
                       pvalueCutoff = .1,
                       TERM2GENE = wpid2gene,
                       TERM2NAME = wpid2name)
ewp.gli.up <- DOSE::setReadable(ewp.gli.up, org.Hs.eg.db, keyType = "ENTREZID")
head(ewp.gli.up)
p1<-barplot(ewp.gli.up, 
            showCategory = 7, 
            font.size = 7,
            label_format = 15)+
  ggtitle("GBM vs LGG DEG up")
ewp.gli.up <- id.table(ewp.gli.up)



#ORA pathway dn
ewp.gli.dn <- enricher(gene = names(gli.dn.gene),
                       universe = names(gli.all.gene),
                       pAdjustMethod = "BH",
                       pvalueCutoff = .1,
                       TERM2GENE = wpid2gene,
                       TERM2NAME = wpid2name)
ewp.gli.dn <- DOSE::setReadable(ewp.gli.dn, org.Hs.eg.db, keyType = "ENTREZID")
head(ewp.gli.dn)
p2<-barplot(ewp.gli.dn,            
            showCategory = 7, 
            font.size = 7,
            label_format = 15)+
  ggtitle("GBM vs LGG DEG down")
ewp.gli.dn <- id.table(ewp.gli.dn,size = 1)

#plot grid for results
pt <- plot_grid(p1, p2,ncol = 2)

## to collect all functional analysis results
fa.DEA2 <- rbind(ewp.gli.up[,keep.colums.enricher],
                 ewp.gli.dn[,keep.colums.enricher])
```

##### GSEA analysis with wikipathways

```{r GSEA in rwikpathways DEA2}
set.seed(1234567)
## GBM vs LGG GSEA
gwp.gli <- GSEA(geneList = c(gli.up.gene,
                             gli.dn.gene),
                pAdjustMethod = "BH",
                pvalueCutoff = 0.1,
                TERM2GENE = wpid2gene,
                TERM2NAME = wpid2name,
                verbose = FALSE
                   )
gwp.gli <- setReadable(gwp.gli, 'org.Hs.eg.db', 'ENTREZID')

#gbm vs lgg gsea wikipathways up
gwp.gli.up <- gwp.gli[which(gwp.gli$NES > 1),]
head(gwp.gli.up)
gwp.gli.up <- id.table(gwp.gli.up, size = 1)

#gbm vs lgg gsea wikipathways down
gwp.gli.dn <- gwp.gli[which(gwp.gli$NES < -1),]
head(gwp.gli.dn)
#gwp.gli.dn <- id.table(gwp.gli.dn) #no results


fa.DEA2 <- rbind(fa.DEA2,
                 setNames(gwp.gli.up[,keep.colums.gsea],names(fa.DEA2)))
# no results     setNames(gwp.gli.dn[,keep.colums.gsea],names(fa.DEA2)))
```

#### Gene Ontology (GO) functional analysis

Gene ontology functional analysis with `clusterProfiler` functions `enrichGo` adn `gseGO`.
Due to the parent-childhood annotation of GO terms, `dropGO` is applied to avoid general terms, eliminating higher levels of GO.

##### ORA in GO

```{r ORA in GO DEA2}
#level to drop
level.go <- 3

### GBM vs LGG
## BIOLOGICAL PROCESS (BP)
# ORA GO up BP
ego.gli.up <- enrichGO(gene = gli.up.genelist$ENSEMBL,
                          OrgDb = org.Hs.eg.db,
                          keyType = 'ENSEMBL',
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.05,
                          readable = TRUE)

ego.gli.up <- setReadable(dropGO(ego.gli.up,level = 1:level.go),
                            org.Hs.eg.db, keyType = "ENTREZID")
head(ego.gli.up)
ego.gli.up <- id.table(ego.gli.up)

# ORA GO down BP
ego.gli.dn <- enrichGO(gene = gli.dn.genelist$ENSEMBL,
                          OrgDb = org.Hs.eg.db,
                          keyType = 'ENSEMBL',
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.05,
                          readable = TRUE)
ego.gli.dn <- setReadable(dropGO(ego.gli.dn,level = 1:level.go),
                            org.Hs.eg.db, keyType = "ENTREZID")
head(ego.gli.dn)
ego.gli.dn <- id.table(ego.gli.dn)



## to collect all functional analysis results 
fa.DEA2 <- rbind(fa.DEA2,
                  ego.gli.up[,keep.colums.enricher], 
                  ego.gli.dn[,keep.colums.enricher])
```


##### GSEA in GO

```{r GSEA in GO DEA2, message=FALSE}
set.seed(1234567)

### GBM vs LGG
## BIOLOGICAL PROCESS (BP)
# GSEA GO BP
ggo.gli.BP <- gseGO(geneList = c(gli.up.gene,
                                 gli.dn.gene),
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pvalueCutoff = 0.05,
                    verbose = FALSE)
ggo.gli.BP <- setReadable(ggo.gli.BP, 'org.Hs.eg.db', 'ENTREZID')

#gbm vs lgg gsea GO up
ggo.gli.up <- ggo.gli.BP[which(ggo.gli.BP$NES > 1),]
head(ggo.gli.up)
ggo.gli.up <- id.table(ggo.gli.up)

#gbm vs lgg gsea GO down
ggo.gli.dn <- ggo.gli.BP[which(ggo.gli.BP$NES < -1),]
head(ggo.gli.dn)
ggo.gli.dn <- id.table(ggo.gli.dn)



## to collect all functional analysis results
fa.DEA2 <- rbind(fa.DEA2,
                  setNames(ggo.gli.up[,keep.colums.gsea],names(fa.DEA2)),
                  setNames(ggo.gli.dn[,keep.colums.gsea],names(fa.DEA2)))
```


#### Kyoto Encyclopedia of Genes and Genomes (KEGG) functional analysis

##### ORA in KEGG

```{r ORA in KEGG DEA2}
### GBM vs LGG
# ORA KEGG up
ekg.gli.up <- enrichKEGG(gene = names(gli.up.gene),
                         organism = "hsa",
                         pAdjustMethod = "BH",
                         universe = names(gli.all.gene),
                         pvalueCutoff = 0.05)

ekg.gli.up <- setReadable(ekg.gli.up, org.Hs.eg.db, keyType = "ENTREZID")
head(ekg.gli.up)
ekg.gli.up <- id.table(ekg.gli.up, size = 16)

# ORA KEGG down
ekg.gli.dn <- enrichKEGG(gene = names(gli.dn.gene),
                         organism = "hsa",
                         pAdjustMethod = "BH",
                         universe = names(gli.all.gene),
                         pvalueCutoff = 0.05)

ekg.gli.dn <- setReadable(ekg.gli.dn, org.Hs.eg.db, keyType = "ENTREZID")
head(ekg.gli.dn)
ekg.gli.dn <- id.table(ekg.gli.dn, size = 2)

## to collect all functional analysis results
fa.DEA2 <- rbind(fa.DEA2,
                  ekg.gli.up[,keep.colums.enricher], 
                  ekg.gli.dn[,keep.colums.enricher])
```


##### GSEA in KEGG


```{r GSEA in KEGG DEA2}
set.seed(1234567)

##GBM vs LGG
#GSEA in KEGG
gkg.gli <- gseKEGG(geneList = c(gli.up.gene,
                                gli.dn.gene),
                   organism = "hsa",
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   verbose = FALSE
                   )
gkg.gli <- setReadable(gkg.gli, 'org.Hs.eg.db', 'ENTREZID')

#gbm vs lgg gsea kegg up
gkg.gli.up <- gkg.gli[which(gkg.gli$NES > 1),]
head(gkg.gli.up)
#gkg.gli.up <- id.table(gkg.gli.up) # no results

#gbm vs lgg gsea kegg down
gkg.gli.dn <- gkg.gli[which(gkg.gli$NES < -1),]
head(gkg.gli.dn)
gkg.gli.dn <- id.table(gkg.gli.dn, size = 1)


fa.DEA2 <- rbind(fa.DEA2,
                 #setNames(gwp.gli.up[,keep.colums.gsea], names(fa.DEA2)),
                 setNames(gkg.gli.dn[,keep.colums.gsea],names(fa.DEA2)))
writexl::write_xlsx(fa.DEA2, path = "DEAsets/fa.DEA2.xlsx")
```

#### Wordcloud functional analysis from DEA2

As the same as analysis 1, in this section different wordclouds are shown from proccess and genes involve in the proccess.

```{r wordclouds FA2}
#### Wordclouds for FA2 ####
### acronyms
## databases
# wp = wikipathways
# go = gene ontology
# kg = KEGG
##cases
# gbm = GBM vs colon
# lgg = LGG vs colon
# gli = GBM vs LGG

### WIKIPATHWAYS ####
## GLIOMAS: GBM vs LGG
#up-regulated
fa.wp.gli.up.2<- proccess.table(fa.table = fa.DEA2, 
                           database = "wp", 
                           cases = "gli", 
                           reg = "up")
head(fa.wp.gli.up.2)
getwordcloud(fa.table = fa.DEA2, 
             database = "wp",
             cases = "gli", 
             reg = "up")
#down-regulated
fa.wp.gli.dn.2<- proccess.table(fa.table = fa.DEA2, 
                           database = "wp", 
                           cases = "gli", 
                           reg = "dn")
head(fa.wp.gli.dn.2)
getwordcloud(fa.table = fa.DEA2, 
             database = "wp",
             cases = "gli", 
             reg = "dn")

## GENE ONTOLOGY ####
## GLIOMAS: GBM vs LGG
#up-regulated
fa.go.gli.up.2<- proccess.table(fa.table = fa.DEA2, 
                           database = "go", 
                           cases = "gli", 
                           reg = "up")
head(fa.go.gli.up.2)
getwordcloud(fa.table = fa.DEA2, 
             database = "go",
             cases = "gli", 
             reg = "up")
#down-regulated
fa.go.gli.dn.2<- proccess.table(fa.table = fa.DEA2, 
                           database = "go", 
                           cases = "gli", 
                           reg = "dn")
head(fa.go.gli.dn.2)
getwordcloud(fa.table = fa.DEA2, 
             database = "go",
             cases = "gli", 
             reg = "dn")

## KEGG ####
## GLIOMAS: GBM vs LGG
#up-regulated
fa.kg.gli.up.2<- proccess.table(fa.table = fa.DEA2, 
                           database = "kg", 
                           cases = "gli", 
                           reg = "up")
head(fa.kg.gli.up.2)
getwordcloud(fa.table = fa.DEA2, 
             database = "kg",
             cases = "gli", 
             reg = "up")
#down-regulated
fa.kg.gli.dn.2<- proccess.table(fa.table = fa.DEA2, 
                           database = "kg", 
                           cases = "gli", 
                           reg = "dn")
head(fa.kg.gli.dn.2)
getwordcloud(fa.table = fa.DEA2, 
             database = "kg",
             cases = "gli", 
             reg = "dn")
```


# Machine Learning analysis

In this section are run several algorithms of machine learning with a classification aim. The algorithms are the following:

From supervised analysis method:  
  + K nearest neighbourhood 
  + Artificial Neural Network 
  + Support Vector Machine
  + Random Forest
  
From non-supervised analysis method:
  + k mean
  + hierarchical clustering
  + Gaussian 

Every algorithms will be run with different gene sets and also contains internal parameter which will be adjusted. Four analysis are done with each algorithm:  
  + All genes from analysis 1: GBM, LGG and colon cases.  
  + DEG for analysis 1: GBM vs colon and LGG vs colon DEA1. 
  + All genes for analysis 2: GBM and LGG cases.  
  + DEG for analysis 2: GBM vs LGG DEA2.  

```{r ML packages, message=FALSE}
library(caret) #evaluation
library(kableExtra) #tables
library(class)#knn
library(mclust) #GBM
library(dbscan)
```


The next chunk is optional to reduce the number of cases for the algorithm execution.
Cases could be reduced due to computational cost. It is optional but it's also recommended for high-cost computational algorithm.

```{r reduce cases for ML algorithm}
##Case reduction for ML analysis is optional
set.seed(987654321)
#optional percentage
percent <- 1

#Cases for FULL data
clust.ids.F <- sort(sample(x =  1:length(colnames(v.F$E)),
                      size = length(colnames(v.F$E))*percent,
                      replace = FALSE))
clust.F <- group_f[clust.ids.F]
prop.table(table(clust.F))
prop.table(table(group_f))

#Cases for SIM data
clust.ids.S <- which(clust.ids.F <= length(colnames(v.S$E)))
clust.S <- group_s[clust.ids.S]

##Assigning number case for patients
#it will be used for plots in ML analysis
# FULL data
patients.id <- c(paste(levels(group_f)[2], 1:length(group_f[which(group_f == "gbm")]),sep = "_"),
                 paste(levels(group_f)[3], 1:length(group_f[which(group_f == "lgg")]),sep = "_"),
                 paste(levels(group_f)[1], 1:length(group_f[which(group_f == "colon")]),sep = "_")
)
colnames(v.F$E) <- patients.id

#SIM data
patients.id <- patients.id[1:length(group_s)]
colnames(v.S$E) <- patients.id
```


Moreover, we need to define some genesets for the ML analysis


```{r gene sets for ML analysis}
#ALL genes from analysis 1
all.a1 <- gbm.all.genelist$ENSEMBL

#Differential expresed genes from DEA1
deg.a1 <- unique(c(gbm.up.genelist$ENSEMBL,
                   gbm.dn.genelist$ENSEMBL,
                   lgg.up.genelist$ENSEMBL,
                   lgg.dn.genelist$ENSEMBL
                   )
                 )
length(deg.a1)

#ALL genes from analysis 2
all.a2 <- gli.all.genelist$ENSEMBL

#Differential expresed genes from DEA2
deg.a2 <- unique(c(gli.up.genelist$ENSEMBL,
                     gli.dn.genelist$ENSEMBL
                     )
                   )
length(deg.a2)
```


```{r function and variables to work with in ML}
#variables to save supervised analysis
reset.var.s <- function(){
  cont.mat <<- data.frame()
  parametro <<- c()
  accuracy <<- c()
  kappa.e <<- c()
}
```

## Machine Learning for analysis 1



```{r variables for analysis 1}
#This chunk save a variable for the supervised analysis.
#col_cont.mat
str_cancer <- c("colon","gbm","lgg")
count <- 1
col_cont.mat <- c()
for (i in 1:3){
  for (j in 1:3){
    col_cont.mat[count] <- paste(str_cancer[i],str_cancer[j], sep = "_as_")
    count <- count + 1
  }
}
```


### Supervised analysis method

For supervised analysis 1 we need to build train and test dataset.

```{r train and test dataset for supervised analysis}
#### Building training and test dataset.
#dataset train and test
set.seed(123456789)

#numbers of data for each dataset
n_train <- round(0.67*length(clust.F),0)
n_test <- round(0.33*length(clust.F),0)

#ids for each dataset
train_ids <- sort(sample(x = 1:length(clust.F),size = n_train,replace = FALSE))
test_ids <- c(1:length(clust.F))[-train_ids]

#building training and test dataset
train_data <- list(all.F = v.F$E[all.a1,train_ids],
                   DEG.F = v.F$E[deg.a1,train_ids])
test_data <- list(all.F = v.F$E[all.a1,test_ids],
                   DEG.F = v.F$E[deg.a1,test_ids])

#labelling dataset
train_label <- clust.F[train_ids]
test_label <- clust.F[test_ids]

#proportions
#train dataset
round(prop.table(table(train_label)),3)*100
#test dataset
round(prop.table(table(test_label)),3)*100
#dataset
round(prop.table(table(clust.F)),3)*100

#short function to extract data train and test
getset <- function(set){
  train <- train_data[[set]]
  test <- test_data[[set]]
  data <- list(train = train,
               test = test)
  return(data)
}
```


#### K nearest neighbourhood (Knn)


```{r knn ML a1 FULL data}
### Knn algorithm
#Evaluating models with different k values (neighbours)
ks <- c(3,5)

#function to evaluate knn models
eval.knn <-function(model, geneset, n.mod,
                      cm = cont.mat,parameter.e = parametro, 
                      accuracy.e = accuracy, kappa = kappa.e){
  modelos <- c(rep(model,n.mod))
  gens <- c(rep(geneset,n.mod))
  resum <- data.frame(modelos, parameter.e, gens,
                    round(accuracy.e,3), round(kappa.e,3), 
                    cont.mat)
  names(resum) <- c("model", "parameter", "geneset", 
                  "accuracy", "kappa", col_cont.mat)
  return(resum)
}

#### DATA all.F ####
## ALL genes analysis FULL data
data <- getset(set = "all.F")

#reseting variables
reset.var.s()

#for loop to execute knn algorithm with differen k values
j <- 0 #counter
for (i in ks){
  j <- j +1 #add 1 to counter
  #execution of knn value k = i
  test_pred <- knn(train = t(data$train), 
                   test = t(data$test), 
                   cl = train_label, 
                   k=i, 
                   prob = TRUE)
  
  #confusion matrix
  conf_knn <- confusionMatrix(test_pred, test_label)
  
  #variables of confusion matrix to save the model
  parametro[j] <- paste("k = ", i)
  cont.mat<- rbind(cont.mat, c(conf_knn[["table"]]))
  accuracy[j] <- conf_knn[["overall"]][["Accuracy"]]
  kappa.e[j] <- conf_knn[["overall"]][["Kappa"]]
}

#evaluation
resum <- eval.knn(model = "kNN",
                  n.mod = 2,
                  geneset = "ALL",
                  parameter.e = parametro,
                  kappa = kappa.e)
#tabla from knn
knn.all.F <- resum
knn.all.F



#### DATA DEG.F ####
## DEG genes analysis FULL data
data <- getset(set = "DEG.F")

#reseting variables
reset.var.s()

#for loop to execute knn algorithm with differen k values
j <- 0 #counter
for (i in ks){
  j <- j +1 #add 1 to counter
  #execution of knn value k = i
  test_pred <- knn(train = t(data$train), 
                   test = t(data$test), 
                   cl = train_label, 
                   k=i, 
                   prob = TRUE)
  
  #confusion matrix
  conf_knn <- confusionMatrix(test_pred, test_label)
  
  #variables of confusion matrix to save the model
  parametro[j] <- paste("k = ", i)
  cont.mat<- rbind(cont.mat, c(conf_knn[["table"]]))
  accuracy[j] <- conf_knn[["overall"]][["Accuracy"]]
  kappa.e[j] <- conf_knn[["overall"]][["Kappa"]]
}

#evaluation
#evaluation for models of knn

resum <- eval.knn(model = "kNN",
                  n.mod = 2,
                  geneset = "DEG",
                  parameter.e = parametro,
                  kappa = kappa.e)
#tabla from knn
knn.DEG.F <- resum
knn.DEG.F
```


### Non-Supervised analysis method

#### K mean

```{r kmean ML a1 FULL data}
### Kmean algorithm
#function to evaluate kmean models
eval.km <- function(result, group.clust,
                    geneset, parameter.e,
                    model = "kmeans"){
  kmeans <- result
  group <- group.clust
                    
  #assign each cluster with the level group
  km_eval <- data.frame(cluster = kmeans,
                        group = group)

  levels_new_group <- c()
  for (i in 1:3){
    km_eval$new_group[km_eval$group == levels(group)[i]] <- 
      which.max(tabulate(km_eval$cluster[km_eval$group == levels(group)[i]]))
  }
  for (i in 1:3){
    levels_new_group[i] <- as.character(km_eval$group[which(km_eval$new_group == i)])[2]
  }
  
  actual_label <- factor(km_eval$group, levels = levels_new_group)
  predict_cluster <- as.factor(km_eval$cluster)
  levels(predict_cluster) <- levels_new_group

  #reseting variables
  reset.var.s()
  
  #to evaluate confusionmatrix
  con_mat <- confusionMatrix(predict_cluster,actual_label)
  cont.mat<- rbind(cont.mat,c(con_mat[["table"]]))
  accuracy <- con_mat[["overall"]][["Accuracy"]]
  kappa.e <- con_mat[["overall"]][["Kappa"]]

  #resumen
  resum <- data.frame(model, parameter.e, geneset,
                    round(accuracy,3), round(kappa.e,3), 
                    cont.mat)
  cont.mat_names <- c()
  for (col in colnames(con_mat[["table"]])){
    for (row in rownames(con_mat[["table"]])){
      cont.mat_names <- append(cont.mat_names, paste(col, "as", row))
    }
  }
  names(resum) <- c("model", "parameter","geneset",
                  "accuracy", "kappa",cont.mat_names)
  return(resum)
}

#### DATA all.F ####
## ALL genes analysis FULL data
data <- t(v.F$E[all.a1, clust.ids.F])

#k mean execution
set.seed(666333666)
km <- kmeans(data,centers = 3) #three groups
fig_clus<- fviz_cluster(object = km,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_clus

##Evaluation


resum <- eval.km(result = km$cluster,
                 group.clust = clust.F,
                 geneset = "ALL",parameter.e = "3 groups")

#kmean table
kme.all.F <- resum
kme.all.F

#### DATA DEG.F ####
## ALL genes analysis FULL data
data <- t(v.F$E[deg.a1, clust.ids.F])

#k mean execution
set.seed(666333666)
km <- kmeans(data,centers = 3) #three groups
fig_clus<- fviz_cluster(object = km,
                        data = data,
                        geom = c("text"),
                        labelsize = 7,
                        ggtheme = theme_bw(),
                        main = paste())
fig_clus

##Evaluation


resum <- eval.km(result = km$cluster,
                 group.clust = clust.F,
                 geneset = "DEG",parameter.e = "3 groups")


#kmean table
kme.DEG.F <- resum
kme.DEG.F
```


#### Hierarchical clustering


```{r hc ML a1 FULL data}
#### Hierarchical clustering

#### DATA all.F ####
## ALL genes analysis FULL data
data <- t(v.F$E[all.a1, clust.ids.F])

set.seed(123)
d <- dist(data, method = "euclidean")
# Hierarchical clustering using Complete Linkage
hc <- hclust(d, method = "ward.D" )
# Cut tree into 3 groups
sub_grp <- cutree(hc, k = 3)

# Plot full dendogram
fig_hc <-fviz_dend(
  hc,
  k = 3,
  horiz = FALSE,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco",
  cex = 0.3
)
fig_hc

#evaluation
#function of evaluation of km model can be used here
resum <- eval.km(result = sub_grp, group.clust = clust.F,
                 geneset = "ALL",parameter.e = "cutree 3",
                 model = "hclust")

hcl.all.F <- resum
hcl.all.F

#### DATA DEG.F ####
## DEG genes analysis FULL data
data <- t(v.F$E[deg.a1, clust.ids.F])

set.seed(123)
d <- dist(data, method = "euclidean")
# Hierarchical clustering using Complete Linkage
hc <- hclust(d, method = "ward.D" )
# Cut tree into 3 groups
sub_grp <- cutree(hc, k = 3)

# Plot full dendogram
fig_hc <-fviz_dend(
  hc,
  k = 3,
  horiz = FALSE,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco",
  cex = 0.3
)
fig_hc

#evaluation
#function of evaluation of km model can be used here
resum <- eval.km(result = sub_grp, group.clust = clust.F,
                 geneset = "DEG",parameter.e = "cutree 3",
                 model = "hclust")

hcl.DEG.F <- resum
hcl.DEG.F
```


#### Gaussian Mixture Model (GMM)

For GMM and all genes dataset algorith couln't be implemented due to stack overflow related problem.

```{r GMM ML a1 FULL data}
#### Gaussian Mixture Model

#### DATA all.F ####
## ALL genes analysis FULL data
data <- t(v.F$E[all.a1, clust.ids.F])

set.seed(1234567)
#gmm model 
#gmm <- Mclust(data, G = 3)

#fig_gmm <- fviz_cluster(object = gmm,
#             data = data,
 #            geom = c("text"),
 #            labelsize = 7,
 #            ggtheme = theme_bw(),
 #            main = paste())
#fig_gmm

#evaluation
#function of evaluation of km model can be  also used here
#resum <- eval.km(result = gmm$classification, group.clust = clust.F,
#                 geneset = "ALL",parameter.e = "G 3",
#                 model = "hclust")
#
#gmm.all.F <- resum
#gmm.all.F

#### DATA DEG.F ####
## DEG genes analysis FULL data
data <- t(v.F$E[deg.a1, clust.ids.F])

set.seed(1234567)
#gmm model 
gmm <- Mclust(data, G = 3)

fig_gmm <- fviz_cluster(object = gmm,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_gmm

#evaluation
#function of evaluation of km model can be  also used here
resum <- eval.km(result = gmm$classification, group.clust = clust.F,
                 geneset = "DEG",parameter.e = "G 3",
                 model = "GMM")

gmm.DEG.F <- resum
gmm.DEG.F
```


#### Density-based spatial clustering of applications with noise (DBSCAN)


```{r DSCAN ML a1 FULL data}
#### DBSCAN clust

#### DATA all.F ####
## ALL genes analysis FULL data
data <- t(v.F$E[all.a1, clust.ids.F])

#to select the variable eps it is displayed a knndistplot
kNNdistplot(data, k = 50)
abline(h = 225, lty = 2)

#DBSCAN algorithm
set_eps <- 225
minpoints <- 25
dbs <- dbscan(data,eps = set_eps, MinPts = minpoints)

fig_dbs <- fviz_cluster(object = dbs,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_dbs

#evaluation
#function of evaluation of km model can be  also used here
resum <- eval.km(result = dbs$cluster, 
                 group.clust = clust.F,
                 geneset = "ALL",parameter.e = "EPS = 225,
                 minp = 25",
                 model = "dbscan")

dbs.all.F <- resum
dbs.all.F

#### DATA DEG.F ####
## DEG genes analysis FULL data
data <- t(v.F$E[deg.a1, clust.ids.F])

#to select the variable eps it is displayed a knndistplot
kNNdistplot(data, k = 25)
abline(h = 160, lty = 2)

#DBSCAN algorithm
set_eps <- 160
minpoints <- 25
dbs <- dbscan(data,eps = set_eps, MinPts = minpoints)

fig_dbs <- fviz_cluster(object = dbs,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_dbs

#evaluation
#function of evaluation of km model can be  also used here
resum <- eval.km(result = dbs$cluster, 
                 group.clust = clust.F,
                 geneset = "DEG",parameter.e = "EPS = 180,
                 minp = 25",
                 model = "dbscan")

dbs.DEG.F <- resum
dbs.DEG.F
```


### Machine Learning conclusions for analysis 1

Chunk for create a table of ML conclusions.

```{r ML conclusions analysis 1}
#final table resum
resum_fin.a1 <- rbind(knn.all.F[,1:5],
                      knn.DEG.F[,1:5],
                      kme.all.F[,1:5],
                      kme.DEG.F[,1:5],
                      hcl.all.F[,1:5],
                      hcl.DEG.F[,1:5],
                      #gmm.all.F[,1:5],
                      gmm.DEG.F[,1:5],
                      dbs.all.F[,1:5],
                      dbs.DEG.F[,1:5]
)

writexl::write_xlsx(x = resum_fin.a1,path = "DEAsets/ml.a1conclusions.xlsx")

kable(resum_fin.a1,row.names = FALSE) %>%
  kable_styling(position = "center")
```



## Machine Learning for analysis 2

```{r variables for analysis 2}
#col_cont.mat
str_cancer <- c("gbm","lgg")
count <- 1
col_cont.mat2 <- c()
for (i in 1:2){
  for (j in 1:2){
    col_cont.mat2[count] <- paste(str_cancer[i],str_cancer[j], sep = "_as_")
    count <- count + 1
  }
}
```

### Supervised analysis method

For supervised analysis 2 we need to build train and test dataset again.

```{r train and test dataset for supervised analysis 2}
#### Building training and test dataset.
#dataset train and test
set.seed(123456789)

#numbers of data for each dataset
n_train <- round(0.67*length(clust.S),0)
n_test <- round(0.33*length(clust.S),0)

#ids for each dataset
train_ids <- sort(sample(x = 1:length(clust.S),size = n_train,replace = FALSE))
test_ids <- c(1:length(clust.S))[-train_ids]

#building training and test dataset
train_data <- list(all.S = v.S$E[all.a2,train_ids],
                   DEG.S = v.S$E[deg.a2,train_ids])
test_data <- list(all.S = v.S$E[all.a2,test_ids],
                   DEG.S = v.S$E[deg.a2,test_ids])

#labelling dataset
train_label <- clust.S[train_ids]
test_label <- clust.S[test_ids]

#proportions
#train dataset
round(prop.table(table(train_label)),3)*100
#test dataset
round(prop.table(table(test_label)),3)*100
#dataset
round(prop.table(table(clust.S)),3)*100
```


#### K nearest neighbourhood (Knn)


```{r knn ML a2 SIM data}
### Knn algorithm
#Evaluating models with different k values (neighbours)
ks <- c(3,5)

#function to evaluate knn models
eval.knn2 <-function(model, geneset, n.mod,
                      cm = cont.mat,parameter.e = parametro, 
                      accuracy.e = accuracy, kappa = kappa.e){
  modelos <- c(rep(model,n.mod))
  gens <- c(rep(geneset,n.mod))
  resum <- data.frame(modelos, parameter.e, gens,
                    round(accuracy.e,3), round(kappa.e,3), 
                    cont.mat)
  names(resum) <- c("model", "parameter", "geneset", 
                  "accuracy", "kappa", col_cont.mat2)
  return(resum)
}

#### DATA all.S ####
## ALL genes analysis sim data
data <- getset(set = "all.S")

#reseting variables
reset.var.s()

#for loop to execute knn algorithm with differen k values
j <- 0 #counter
for (i in ks){
  j <- j +1 #add 1 to counter
  #execution of knn value k = i
  test_pred <- knn(train = t(data$train), 
                   test = t(data$test), 
                   cl = train_label, 
                   k=i, 
                   prob = TRUE)
  
  #confusion matrix
  conf_knn <- confusionMatrix(test_pred, test_label)
  
  #variables of confusion matrix to save the model
  parametro[j] <- paste("k = ", i)
  cont.mat<- rbind(cont.mat, c(conf_knn[["table"]]))
  accuracy[j] <- conf_knn[["overall"]][["Accuracy"]]
  kappa.e[j] <- conf_knn[["overall"]][["Kappa"]]
}

#evaluation
resum <- eval.knn2(model = "kNN",
                  n.mod = 2,
                  geneset = "ALL",
                  parameter.e = parametro,
                  kappa = kappa.e)
#tabla from knn
knn.all.S <- resum
knn.all.S



#### DATA DEG.S ####
## DEG genes analysis sim data
data <- getset(set = "DEG.S")

#reseting variables
reset.var.s()

#for loop to execute knn algorithm with differen k values
j <- 0 #counter
for (i in ks){
  j <- j +1 #add 1 to counter
  #execution of knn value k = i
  test_pred <- knn(train = t(data$train), 
                   test = t(data$test), 
                   cl = train_label, 
                   k=i, 
                   prob = TRUE)
  
  #confusion matrix
  conf_knn <- confusionMatrix(test_pred, test_label)
  
  #variables of confusion matrix to save the model
  parametro[j] <- paste("k = ", i)
  cont.mat<- rbind(cont.mat, c(conf_knn[["table"]]))
  accuracy[j] <- conf_knn[["overall"]][["Accuracy"]]
  kappa.e[j] <- conf_knn[["overall"]][["Kappa"]]
}

#evaluation
#evaluation for models of knn

resum <- eval.knn2(model = "kNN",
                  n.mod = 2,
                  geneset = "DEG",
                  parameter.e = parametro,
                  kappa = kappa.e)
#tabla from knn
knn.DEG.S <- resum
knn.DEG.S
```


### Non-Supervised analysis method


#### K mean


```{r kmean ML a2 sim data}
### Kmean algorithm
#function to evaluate kmean models
eval.km2 <- function(result, group.clust,
                    geneset, parameter.e,
                    model = "kmeans"){
  kmeans <- result
  group <- group.clust
                    
  #assign each cluster with the level group
  km_eval <- data.frame(cluster = kmeans,
                        group = group)

  levels_new_group <- c()
  for (i in 1:2){
    km_eval$new_group[km_eval$group == levels(group)[i]] <- 
      which.max(tabulate(km_eval$cluster[km_eval$group == levels(group)[i]]))
  }
  for (i in 1:2){
    levels_new_group[i] <- as.character(km_eval$group[which(km_eval$new_group == i)])[2]
  }
  
  actual_label <- factor(km_eval$group, levels = levels_new_group)
  predict_cluster <- as.factor(km_eval$cluster)
  levels(predict_cluster) <- levels_new_group

  #reseting variables
  reset.var.s()
  
  #to evaluate confusionmatrix
  con_mat <- confusionMatrix(predict_cluster,actual_label)
  cont.mat<- rbind(cont.mat,c(con_mat[["table"]]))
  accuracy <- con_mat[["overall"]][["Accuracy"]]
  kappa.e <- con_mat[["overall"]][["Kappa"]]

  #resumen
  resum <- data.frame(model, parameter.e, geneset,
                    round(accuracy,3), round(kappa.e,3), 
                    cont.mat)
  cont.mat_names <- c()
  for (col in colnames(con_mat[["table"]])){
    for (row in rownames(con_mat[["table"]])){
      cont.mat_names <- append(cont.mat_names, paste(col, "as", row))
    }
  }
  names(resum) <- c("model", "parameter","geneset",
                  "accuracy", "kappa",cont.mat_names)
  return(resum)
}

#### DATA all.S ####
## ALL genes analysis FULL data
data <- t(v.S$E[all.a2, clust.ids.S])

#k mean execution
set.seed(1234567)
km <- kmeans(data,centers = 2) #two groups
fig_clus<- fviz_cluster(object = km,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_clus

##Evaluation


resum <- eval.km(result = km$cluster,
                 group.clust = clust.S,
                 geneset = "ALL",parameter.e = "2 groups")

#kmean table
kme.all.S <- resum
kme.all.S

#### DATA DEG.F ####
## ALL genes analysis FULL data
data <- t(v.S$E[deg.a2, clust.ids.S])

#k mean execution
#set.seed(1234)
km <- kmeans(data,centers = 2) #three groups
fig_clus<- fviz_cluster(object = km,
                        data = data,
                        geom = c("text"),
                        labelsize = 7,
                        ggtheme = theme_bw(),
                        main = paste())
fig_clus

##Evaluation


resum <- eval.km(result = km$cluster,
                 group.clust = clust.S,
                 geneset = "DEG",parameter.e = "2 groups")


#kmean table
kme.DEG.S <- resum
kme.DEG.S
```


#### Hierarchical clustering


```{r hc ML a2 sim data}
#### Hierarchical clustering

#### DATA all.S ####
## ALL genes analysis sim data
data <- t(v.S$E[all.a2, clust.ids.S])

set.seed(123)
d <- dist(data, method = "euclidean")
# Hierarchical clustering using Complete Linkage
hc <- hclust(d, method = "ward.D" )
# Cut tree into 2 groups
sub_grp <- cutree(hc, k = 2)

# Plot full dendogram
fig_hc <-fviz_dend(
  hc,
  k = 2,
  horiz = FALSE,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco",
  cex = 0.3
)
fig_hc

#evaluation
#function of evaluation of km model can be used here
resum <- eval.km(result = sub_grp, group.clust = clust.S,
                 geneset = "ALL",parameter.e = "cutree 2",
                 model = "hclust")

hcl.all.S <- resum
hcl.all.S

#### DATA DEG.S ####
## DEG genes analysis sim data
data <- t(v.S$E[deg.a2, clust.ids.S])

set.seed(123)
d <- dist(data, method = "euclidean")
# Hierarchical clustering using Complete Linkage
hc <- hclust(d, method = "ward.D" )
# Cut tree into 2 groups
sub_grp <- cutree(hc, k = 2)

# Plot full dendogram
fig_hc <-fviz_dend(
  hc,
  k = 2,
  horiz = FALSE,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco",
  cex = 0.3
)
fig_hc

#evaluation
#function of evaluation of km model can be used here
resum <- eval.km(result = sub_grp, group.clust = clust.S,
                 geneset = "DEG",parameter.e = "cutree 2",
                 model = "hclust")

hcl.DEG.S <- resum
hcl.DEG.S
```


#### Gaussian Mixture Model (GMM)

For GMM and all genes dataset algorith couln't be implemented due to stack overflow related problem.

```{r GMM ML a2 sim data}
#### Gaussian Mixture Model

#### DATA all.S ####
## ALL genes analysis sim data
data <- t(v.S$E[all.a2, clust.ids.S])

set.seed(1234567)
#gmm model 
#gmm <- Mclust(data, G = 2)

#fig_gmm <- fviz_cluster(object = gmm,
#             data = data,
#             geom = c("text"),
#             labelsize = 7,
#             ggtheme = theme_bw(),
#             main = paste())
#fig_gmm

#evaluation
#function of evaluation of km model can be  also used here
#resum <- eval.km(result = gmm$classification, group.clust = clust.S,
#                 geneset = "ALL",parameter.e = "G 2",
#                 model = "GMM")

#gmm.all.S <- resum
#gmm.all.S

#### DATA DEG.S ####
## DEG genes analysis sim data
data <- t(v.S$E[deg.a2, clust.ids.S])

set.seed(1234567)
#gmm model 
gmm <- Mclust(data, G = 2)

fig_gmm <- fviz_cluster(object = gmm,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_gmm

#evaluation
#function of evaluation of km model can be  also used here
resum <- eval.km(result = gmm$classification, group.clust = clust.S,
                 geneset = "DEG",parameter.e = "G 2",
                 model = "GMM")

gmm.DEG.S <- resum
gmm.DEG.S
```


#### Density-based spatial clustering of applications with noise (DBSCAN)


```{r DSCAN ML a2 sim data}
#### DBSCAN clust

#### DATA all.S ####
## ALL genes analysis sim data
data <- t(v.S$E[all.a2, clust.ids.S])

#to select the variable eps it is displayed a knndistplot
kNNdistplot(data, k = 15)
abline(h = 150, lty = 2)

#DBSCAN algorithm
set_eps <- 140
minpoints <- 15
dbs <- dbscan(data,eps = set_eps, MinPts = minpoints)

fig_dbs <- fviz_cluster(object = dbs,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_dbs #only one group so it won't be evaluated

#evaluation
#function of evaluation of km model can be  also used here
resum <- eval.km(result = dbs$cluster, 
                 group.clust = clust.S,
                 geneset = "ALL",parameter.e = "EPS = 140,
                 minp = 15",
                 model = "dbscan")

dbs.all.S <- resum
dbs.all.S

#### DATA DEG.S ####
## DEG genes analysis sim data
data <- t(v.S$E[deg.a2, clust.ids.S])

#to select the variable eps it is displayed a knndistplot
kNNdistplot(data, k = 15)
abline(h = 55, lty = 2)

#DBSCAN algorithm
set_eps <- 50
minpoints <- 15
dbs <- dbscan(data,eps = set_eps, MinPts = minpoints)

fig_dbs <- fviz_cluster(object = dbs,
             data = data,
             geom = c("text"),
             labelsize = 7,
             ggtheme = theme_bw(),
             main = paste())
fig_dbs

#evaluation
#function of evaluation of km model can be  also used here
resum <- eval.km(result = dbs$cluster, 
                 group.clust = clust.S,
                 geneset = "DEG",parameter.e = "EPS = 50,
                 minp = 15",
                 model = "dbscan")

dbs.DEG.S <- resum
dbs.DEG.S
```

### Machine Learning conclusions for analysis 2

Chunk for create a table of ML conclusions.

```{r ML conclusions analysis 2}
#final table resum
resum_fin.a2 <- rbind(knn.all.S[,1:5],
                      knn.DEG.S[,1:5],
                      kme.all.S[,1:5],
                      kme.DEG.S[,1:5],
                      hcl.all.S[,1:5],
                      hcl.DEG.S[,1:5],
                      #gmm.all.S[,1:5],
                      gmm.DEG.S[,1:5],
                      dbs.all.S[,1:5],
                      dbs.DEG.S[,1:5])


kable(resum_fin.a2,row.names = FALSE) %>%
  kable_styling(position = "center")
```

## References




